---
title: "Fuel efficiency of manual and automatic transmissions"
author: "Andrea Schioppa"
output: html_document
---

```{r loading, eval = TRUE, echo = FALSE}
data(mtcars)
library(xtable)
fit <- lm(mpg ~ am * wt + hp, data = mtcars)
options(xtable.comment=FALSE)
```

```{r new_fit, eval=TRUE, echo = FALSE}
newmtcars <- mtcars[-c(17, 20, 31, 18),]
fit.new <- lm(mpg ~ am * wt + hp, data = newmtcars)
cf.new <- coef(fit.new)
sfit.new <- summary(fit.new)
```

```{r am_conf, eval = TRUE, echo = FALSE}
interval_am <- sfit.new$coefficients[2,1] + c(-1,1) * sfit.new$coefficients[2,2] *
    qt(0.975, df = sfit.new$df[2])
```

# Summary

We use the dataset `mtcars` to compare the fuel efficiency of automatic and manual trasmissions.
We fit a linear model predicting the miles per US gallon (`mpg`) using the transmission (`am=0` for
automatic, `am=1` for manual), the weight `wt` (in 1000lbs) and the gross horsepower `hp` as predictors;
the choice of the model is motivated by analysis of variance.

We find that manual transmission leads to an estimated increase of
`mpg` of `r sprintf("%1.2f",cf.new[2])` with a $95\%$ confidence interval  [`r sprintf("%1.2f",interval_am[1])`,`r sprintf("%2.2f",interval_am[2])`].

However, we also find that with increasing weight automatic cars lose
less efficiency (see interpretation of the coefficient `am:wt`).
A limitation of the model is that manual cars tend to weigh less than automatic ones, so comparing efficiency
for a given weight seems reasonable in a range of about $[2500,3500]$ lbs.

# Analysis

The data set `mtcars` consists of 32 observations of 11 variables. In Figure 1 we make some exploratory
plots to compare `mpg` with the other variables. Besides using `am` (transmission type), we
decide to try as continuous predictors `wt`, `hp`, `disp`,  and as categorical predictors
`cyl`, `vs`, `gear`.

To select the model we first tried

```{r selection_continuous, message=FALSE}
fit1 <- lm(mpg ~ am + wt, data=mtcars); fit2 <- update(fit1, mpg ~ am*wt)
fit3 <- update(fit2, . ~ .+ hp); fit4 <- update(fit3, . ~ . + am*hp);
anova(fit1,fit2,fit3,fit4)$"Pr(>F)"
```

because of the p-values we decide to use the model `mpg~am*wt+hp`. We have also tried to include any of `disp`,
`vs`, `cyl` and `gear`: `anova` suggests that adding any of these predictors is not significant (we
omit the code for reasons of space).

We fit the model with `fit <- lm(mpg ~ am * wt + hp, data = mtcars)`; the diagnostic graphs
in Figures 2,3 indicate "outliers", Maserati Bora(31), Chrysler Imperial(17), Toyota Corolla(20), Fiat 128(18)
that we decide to remove after inspecting `dfbetas(fit)`, `hatvalues(fit)`.

We refit the model and observe an improvement through the quantile-quantile plot Figure 4.
```{r, ref.label = "new_fit", eval = FALSE}
```

We show a summary of the coefficients and p-values.
```{r summary, echo=FALSE, results="asis"}
sum.table <- sfit.new$coefficients[,c(-3)]
colnames(sum.table)[3] <- "p-value"
xtable(sum.table, display=c("s", "f", "e", "e"))
```

The coefficient `am` is interpreted as the expected increase in `mpg` when considering
a manual vs. an automatic transmission. The coefficient `wt` is interpreted as the
change in `mpg` when increasing the weight by 1 unit (=1000 lbs) for an automatic transmission.
Finally, `am:wt` quantifies the change in `wt` if we consider a manual transmission.

The 95% confidence interval for `am` is:

```{r, ref.label =  "am_conf", eval=FALSE}
```
```{r echoing am_conf}
interval_am
```

We similarly compute the 95% confidence interval for `wt`:

```{r wt_conf, echo=FALSE}
interval_wt_auto <- sfit.new$coefficients[3,1] + c(-1,1) * sfit.new$coefficients[3,2] *
    qt(0.975, df = sfit.new$df[2])
interval_wt_auto
```
and `am:wt`:

```{r wt_ratio, echo=FALSE}
interval_wt_manual_ratio <- sfit.new$coefficients[5,1] + c(-1,1) * sfit.new$coefficients[5,2] *
    qt(0.975, df = sfit.new$df[2])
interval_wt_manual_ratio
```

# Appendix: Figures

```{r exploratory_plot1, echo=FALSE, dev = 'pdf', fig.height=4, fig.cap='Exploratory plot 1'}
par(mfrow=c(2,5), mar=c(4,4,2,1), oma=c(0,0,2,0))
with(mtcars, plot(mpg~cyl, type = "n", main = "mpg vs. cyl"))
with(subset(mtcars, am==0), points(cyl, mpg, col="blue"))
with(subset(mtcars, am==1), points(cyl, mpg, col="red"))
legend("topright", pch=1, col=c("blue", "red"), legend = c("automatic", "manual"), cex=0.8)
with(mtcars, plot(mpg~hp, type = "n", main = "mpg vs. hp"))
with(subset(mtcars, am==0), points(hp, mpg, col="blue"))
with(subset(mtcars, am==1), points(hp, mpg, col="red"))
with(mtcars, plot(mpg~wt, type = "n", main = "mpg vs. wt"))
with(subset(mtcars, am==0), points(wt, mpg, col="blue"))
with(subset(mtcars, am==1), points(wt, mpg, col="red"))
with(mtcars, plot(mpg~qsec, type = "n", main = "mpg vs. qsec"))
with(subset(mtcars, am==0), points(qsec, mpg, col="blue"))
with(subset(mtcars, am==1), points(qsec, mpg, col="red"))
with(mtcars, plot(mpg~drat, type = "n", main = "mpg vs. drat"))
with(subset(mtcars, am==0), points(drat, mpg, col="blue"))
with(subset(mtcars, am==1), points(drat, mpg, col="red"))
with(mtcars, plot(mpg~gear, type = "n", main = "mpg vs. gear"))
with(subset(mtcars, am==0), points(gear, mpg, col="blue"))
with(subset(mtcars, am==1), points(gear, mpg, col="red"))
with(mtcars, plot(mpg~am, type = "n", main = "mpg vs. am"))
with(subset(mtcars, am==0), points(am, mpg, col="blue"))
with(subset(mtcars, am==1), points(am, mpg, col="red"))
with(mtcars, plot(mpg~vs, type = "n", main = "mpg vs. vs"))
with(subset(mtcars, am==0), points(vs, mpg, col="blue"))
with(subset(mtcars, am==1), points(vs, mpg, col="red"))
with(mtcars, plot(mpg~disp, type = "n", main = "mpg vs. disp"))
with(subset(mtcars, am==0), points(disp, mpg, col="blue"))
with(subset(mtcars, am==1), points(disp, mpg, col="red"))
with(mtcars, plot(mpg~carb, type = "n", main = "mpg vs. carb"))
with(subset(mtcars, am==0), points(carb, mpg, col="blue"))
with(subset(mtcars, am==1), points(carb, mpg, col="red"))
```

```{r diagnostic_plot1, echo=FALSE, dev = 'pdf', fig.height=5, fig.cap='Diagnostic plot 1'}
par(mfrow = c(1,2))
plot(fit, which=1)
plot(fit, which=2)
```

```{r diagnostic_plot2, echo=FALSE, dev = 'pdf', fig.height=5, fig.cap='Diagnostic plot 2'}
plot(fit, which=5)
```

```{r qqplot, echo=FALSE, dev = 'pdf', fig.height=5, fig.cap='QQ-plot after removing "outliers"'}
plot(fit.new, which=2)
```